Identifying Effective Motivational Interviewing Sequences Using Automated Pattern Analysis

### 2018-06-22 jhir/latex-manuscript/manuscript.tex
incorporate co-authors feedback

This study is part of a line of research to develop machine-learning models to annotate (code) and analyze patient-counselor communication patterns. We have previously reported on the development of probabilistic generative models~\cite{kotov2014towards, kotov2015interpretable} and application of novel features for maximum margin and deep learning classifiers~\cite{hasan2016study} with the goal of automated annotation of MI session transcripts. Experiments applying the annotation model to novel datasets are underway to assess the generalizability of the model to more diverse types of clinical encounters (e.g, email coaching to increase fruit and vegetable intake, HIV clinical care visits~\cite{carcone2018using}). We also developed and evaluated probabilistic and deep learning methods for the task of predicting success of motivational interviews at any point during the interview~\cite{hasan2018predicting}. \hl{In this paper, we present the current work builds on past work performed on automation of annotation, more specifically, the sequential analysis of patient-counselor communication data to identify the counselor communication strategies linked to the elicitation of change talk and sustain talk.} Next steps include examining the performance of these models in diverse data sets representing different populations and behavioral problems. We are also experimenting with machine learning methods to segment (parse) the stream of communications in healthy eating promotion intervention delivered by email into semantically coherent chunks (phrases, sentences or paragraphs), to which the annotation model can then assign a behavior code. Together, these models form the basis of a complete system to automatically code and analyze patient-counselor interactions. An automated system for behavioral coding and analysis could substantially accelerate the pace of research on the causal mechanisms of Motivational Interviewing and inform both the theory and clinical practice by providing clinicians with information about how to best tailor their communication strategies to different patient populations.

This study is limited by the use of one dataset composed of 37 Motivational Interviewing transcripts of counseling sessions with African American adolescents in weight loss treatment. Thus, there is a need to replicate these findings with larger and more diverse data samples as the findings may not be representative of communication patterns in other contexts employing the Motivational Interviewing framework. In fact, when interpreted in light of the published literature, the results obtained in these experiments suggest that communication patterns are likely to vary given the treatment context. There are, however, consistencies with previous Motivational Interviewing process studies providing support for the validity of our findings and suggesting some counselor communication strategies may cut across treatment contexts. Next steps for this work include enhancing the performance and utility of machine learning models for sequential analysis by combining transcript annotation with sequential analysis. Another limitation of this work is the fact that successful and unsuccessful sequences were analyzed independently. One implication of this approach is that the utility of a counselor behavior, such as the provision of information, to shift an interaction destined for unsuccess to success, cannot be determined from these analyses.

>>>>

This study is part of a line of research to develop machine-learning models to annotate (code) and analyze patient-counselor communication patterns. We have previously reported on the development of probabilistic generative models~\cite{kotov2014towards, kotov2015interpretable} and application of novel features for maximum margin and deep learning classifiers~\cite{hasan2016study} with the goal of automated annotation of MI session transcripts. Experiments applying the annotation model to novel datasets are underway to assess the generalizability of the model to more diverse types of clinical encounters (e.g, email coaching to increase fruit and vegetable intake, HIV clinical care visits~\cite{carcone2018using}). We also developed and evaluated probabilistic and deep learning methods for the task of predicting success of motivational interviews at any point during the interview~\cite{hasan2018predicting}. \hl{In this paper, we present current work that builds on past work performed on automation of annotation, more specifically, the sequential analysis of patient-counselor communication data to identify the counselor communication strategies linked to the elicitation of change talk and sustain talk.} Next steps include examining the performance of these models in diverse data sets representing different populations and behavioral problems. We are also experimenting with machine learning methods to segment (parse) the stream of communications \hl{in a healthy} eating promotion intervention delivered by email into semantically coherent chunks (phrases, sentences or paragraphs), to which the annotation model can then assign a behavior code. Together, these models form the basis of a complete system to automatically code and analyze patient-counselor interactions. An automated system for behavioral coding and analysis could substantially accelerate the pace of research on the causal mechanisms of Motivational Interviewing and inform both the theory and clinical practice by providing clinicians with information about how to best tailor their communication strategies to different patient populations.

This study is limited by the use of one dataset composed of 37 Motivational Interviewing transcripts of counseling sessions with African American adolescents in weight loss treatment. Thus, there is a need to replicate these findings with larger and more diverse data samples as the findings may not be representative of communication patterns in other contexts employing the Motivational Interviewing framework. In fact, when interpreted in light of the published literature, the results obtained in these experiments suggest that communication patterns are likely to vary given the treatment context. There are, however, consistencies with previous Motivational Interviewing process studies providing support for the validity of our findings and suggesting some counselor communication strategies may cut across treatment contexts. Next steps for this work include enhancing the performance and utility of machine learning models for sequential analysis by combining transcript annotation with sequential analysis. Another limitation of this work is the fact that successful and unsuccessful sequences were analyzed independently. One implication of this approach is that the utility of a counselor behavior, such as the provision of information, to shift an interaction destined \hl{for failure to success}, cannot be determined from these analyses.
